Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-10-15 analysis/paper/paper.Rmd
submission prep in progress

NOTE: please provide comments and suggestions for the introduction here:
<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>.

@@

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
Various efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that uncertainty in models
informed only by local demographic data could be used to delay conservation action for poorly monitored populations. We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships [@johnson_science_2020] with
available local demographic data in a simple Bayesian population model
for boreal caribou.

@@

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, can improve viability projections on changing
landscapes, and can help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach will help to identify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

@@

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$.

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females ($\dot{W}_t$)
and the maximum
potential recruitment of female calves per cow ($\dot{X}_t$):

@@

NOTE: Please provide comments and suggestions for the discussion here:
<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.

@@

We have proposed a simple Bayesian population model that helps address these challenges by
acknowledging uncertainty associated with local demographic data and
national demographic-disturbance relationships, and integrating these
two sources of information to reduce uncertainty in projections of the
impacts of changing landscape conditions. This Bayesian approach
represents an important improvement over less formally integrated
methods for estimating and projecting impacts of disturbance on
population population viability [@eccc_scientific_2011;
@johnson_science_2020] because it reconciles and accounts for
uncertainties in local demographic data and in demographic-disturbance
relationships.

Our model assumes that increasing cumulative anthropogenic disturbance
causes a log-linear decrease in caribou demographic rates (equations 4,
5, 10 and 13), reflecting a scientific consensus that cumulative anthropogenic disturbance has a
negative impact on boreal caribou [@eccc_scientific_2011;
@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
@rudolph_demographic_2017; @sorensen_determining_2008;
@hervieux_widespread_2013]. Local data can be integrated to estimate
local demographic-disturbance parameters that differ from the national
means, and in the absence of local data, appropriate priors lead to a
range of outcomes consistent with the observed range of variation across
the country [Figures 3 and 5 in @johnson_science_2020]. Others
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @dalgarno_bboutools_2024] have also used Bayesian models to estimate
caribou demographic rates from local monitoring data, but our addition
of prior information from national demographic-disturbance relationships
is novel, and important for reducing the risk that lack of local data
will be used as a reason to delay conservation action for highly
disturbed and poorly monitored populations. Integrating
demographic-disturbance relationships into a viability projection model
clarifies that no additional information is needed to project outcomes
where disturbance is high and actions to support recovery are not taken.

>>>>

@@

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
Various efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@bissonnette_controverses_2020; @boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that uncertainty in models
informed only by local demographic data could be used to delay conservation action for poorly monitored populations. We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships [@johnson_science_2020] with
available local demographic data in a simple Bayesian population model
for boreal caribou.

@@

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, can improve viability projections on changing
landscapes, and can help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach will help to identify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018].

@@

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Realized recruitment rate varies with population density [@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females ($\dot{W}_t$)
and the maximum
potential recruitment of female calves per cow ($\dot{X}_t$):

@@

@@

We have proposed a simple Bayesian population model that helps address these challenges by
acknowledging uncertainty associated with local demographic data and
national demographic-disturbance relationships, and integrating these
two sources of information to reduce uncertainty in projections of the
impacts of changing landscape conditions. This Bayesian approach
represents an important improvement over less formally integrated
methods for estimating and projecting impacts of disturbance on
population caribou population viability [@eccc_scientific_2011;
@johnson_science_2020] because it reconciles and accounts for
uncertainties in local demographic data and in demographic-disturbance
relationships.

Our model assumes that increasing cumulative anthropogenic disturbance
causes a log-linear decrease in caribou demographic rates (equations 4,
5, 10 and 13), reflecting a scientific consensus that cumulative anthropogenic disturbance has a
negative impact on boreal caribou [@eccc_scientific_2011;
@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
@rudolph_demographic_2017; @sorensen_determining_2008;
@hervieux_widespread_2013]. Local data can be integrated to estimate
local demographic-disturbance parameters that differ from the national
means, and in the absence of local data, appropriate priors lead to a
range of outcomes consistent with the observed range of variation across
the country [Figures 3 and 5 in @johnson_science_2020]. Others
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @dalgarno_bboutools_2024] have also used Bayesian models to estimate
caribou demographic rates from local monitoring data, but our addition
of prior information from national demographic-disturbance relationships
is novel, and important for reducing the risk that lack of local data
will be used as a reason to delay conservation action for highly
disturbed and poorly monitored populations [@bissonnette_controverses_2020; @boan_climate_2018]. Integrating
demographic-disturbance relationships into a viability projection model
clarifies that no additional information is needed to project outcomes
where disturbance is high and actions to support recovery are not taken.
